Optimal forward modeling of gravitational fields using tesseroids with adaptive discretization

### 2014-09-10 body.tex
Dist/size ratio should be D, r is radius already

\begin{equation}
    \frac{d}{L} > r,
    \label{eq:ratio}
\end{equation}

\noindent
for a positive non-zero scalar $r$,
henceforth referred to as
the ``distance-size ratio''.

@@

\begin{verbatim}
function gravity(P, model):
  result = 0
  for each tesseroid in model:
    d = distance from P to the top of tesseroid
    if d/L_lon < r:
      divide_lon = true
    if d/L_lat < r:
      divide_lat = true
    if d/L_r < r:
      divide_r = true
    if divide_lon or divide_lat or divide_r:
      newmodel = (divide tesseroid in the
                  required dimensions)
      result = result + gravity(P, newmodel)
    else
      result = result + effect of tesseroid
  return result
\end{verbatim}

The distance-size ratio $r$ controls
how far away $P$ has to be
compared to the dimensions of the tesseroid.
A ratio $r=2$ means that
the recursive division algorithm will ensure that
tesseroids are smaller than
half the distance to $P$.
Thus, the ratio determines how many divisions are required.
Fig.~\ref{fig:ratio}
illustrates the recursive division of a tesseroid
(Fig.~\ref{fig:ratio}a)
using increasing values of $r$.
The tesseroid is divided into
$28$ tesseroids for $r=1$ (Fig.~\ref{fig:ratio}b),
$146$ tesseroids for $r=2$ (Fig.~\ref{fig:ratio}c),
and $3106$ tesseroids for $r=6$ (Fig.~\ref{fig:ratio}d).

Because the number of tesseroids increases with $r$,
so does the computation time.
Hence, there is need to determine
an optimal value of the distance-size ratio $r$
that yields an acceptable trade-off between
accuracy and computation time.
In the following section,
we describe a methodology
to quantify the error of the GLQ integration
as a function of the distance-size ratio $r$.
These results can be used
to determine optimal values of $r$
for the gravitational potential,
gravitational attraction,
and gravity gradient tensor.

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $r$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $r$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:ratio}
\end{figure}

@@

\noindent
where $r$ is the radial coordinate of the computation point and
$r_1 = R$ and $r_2 = r_1 + \Delta r$ are radial coordinates
of the bottom and top of the spherical shell, respectively.
Table~\ref{tab:shell} summarises the parameters defining the spherical shell.

The computation point can be considered at any longitude or latitude because of
the symmetry of the spherical shell.

The spherical shell had dimensions of $r_1 = R$ ($R$ is the mean Earth radius) and $r_2 = R +
1\ km$ and density $\rho = 2.67\ g.cm^{-3}$.

The largest errors due are concentrated directly on top of the tesseroid
(Fig.~\ref{fig:sample}b).

Therefore, we perform the computations on an $11 \times 11$ point regular grid
in an area directly on top of the tesseroid with corner at $\lambda = 0^\circ$
$\phi = 0^\circ$.

The computation grid was located at a constant height above the spherical shell
$h = r_2 + 1\ km$.

>>>>

\begin{equation}
    \frac{d}{L} > D,
    \label{eq:ratio}
\end{equation}

\noindent
for a positive non-zero scalar $D$,
henceforth referred to as
the ``distance-size ratio''.

@@

\begin{verbatim}
function gravity(P, model):
  result = 0
  for each tesseroid in model:
    d = distance from P to the top of tesseroid
    if d/L_lon < D:
      divide_lon = true
    if d/L_lat < D:
      divide_lat = true
    if d/L_r < D:
      divide_r = true
    if divide_lon or divide_lat or divide_r:
      newmodel = (divide tesseroid in the
                  required dimensions)
      result = result + gravity(P, newmodel)
    else
      result = result + effect of tesseroid
  return result
\end{verbatim}

The distance-size ratio $D$ controls
how far away $P$ has to be
compared to the dimensions of the tesseroid.
A ratio $D=2$ means that
the recursive division algorithm will ensure that
tesseroids are smaller than
half the distance to $P$.
Thus, the ratio determines how many divisions are required.
Fig.~\ref{fig:ratio}
illustrates the recursive division of a tesseroid
(Fig.~\ref{fig:ratio}a)
using increasing values of $D$.
The tesseroid is divided into
$28$ tesseroids for $D=1$ (Fig.~\ref{fig:ratio}b),
$146$ tesseroids for $D=2$ (Fig.~\ref{fig:ratio}c),
and $3106$ tesseroids for $D=6$ (Fig.~\ref{fig:ratio}d).

Because the number of tesseroids increases with $D$,
so does the computation time.
Hence, there is need to determine
an optimal value of the distance-size ratio $D$
that yields an acceptable trade-off between
accuracy and computation time.
In the following section,
we describe a methodology
to quantify the error of the GLQ integration
as a function of the distance-size ratio $D$.
These results can be used
to determine optimal values of $D$
for the gravitational potential,
gravitational attraction,
and gravity gradient tensor.

\begin{figure}
    \centering
    \includegraphics{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using the distance-size ratio $D$ equal to
        (b) 1, (c) 2, and (d) 6.
        $d$ is the Cartesian distance between
        the center of the top face of the tesseroid
        and $P$.
        $L_r$, $L_\phi$, and $L_\lambda$ are the dimensions of the tesseroid.
        Note that increasing $D$
        results in a fine division of the tesseroid
        close the computation point
        and a coarser division further away.
    }
    \label{fig:ratio}
\end{figure}

@@

\noindent
where $r$ is the radial coordinate of the computation point and
$r_1$ and $r_2$ are radial coordinates
of the bottom and top of the spherical shell, respectively.
Table~\ref{tab:shell} summarizes the parameters defining the spherical shell.

The shell was discretized into tesseroids with $S \times S \times 1\ km$ size.

We then compared the result of the spherical shell with that calculated for the
tesseroid model.

The tesseroid calculations were performed with the recursive division
algorithm.

The tesseroid calculations were repeated for values of the distance/size ratio
$D$ ranging from zero (i.e., no divisions) to eight with an interval of 0.5.

The computation point can be considered at any longitude and latitude because of
the symmetry of the spherical shell.

Fig.~\ref{fig:sample}b shows that the largest errors are concentrated directly
on top of the tesseroid.

Making the comparison between the tesseroid

We perform the computations on an $11 \times 11$ point regular grid.

The southwestern corner of the grid is $(\lambda, \phi) = (0, 0)$
and the northeastern corner is $(S, S)$, i.e., directly on top of a tesseroid.

The grid is located at a constant height $h = 1\ km$ above the spherical shell.

The computations where repeated for discretization into tesseroids of sizes $S = 1^\circ$, $2^\circ$,
$5^\circ$, and $30^\circ$.
